---
title: "An Introduction to multi-factors"
author: "Jason Thorpe [jdthorpe_at_gmail_dot_com]"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    self_contained: no
vignette: >
  %\VignetteIndexEntry{An Introduction to multi-factors}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Why mfactors?

Multi-factors came about in order to deal with character vectors whose
contents are delimited fields, whose entries *should* be limited to a fixed
set of values (levels), and whose values *should* be limited to a single
value per entry.   If we we remove the words 'delimited' and '*should*' from
the prevous sentance, we would have a good definition of a `factor`
variable.  

However, the process that generates such data may not be constrained.
Specifically the data that motivated this class were derives from paper
surveys in which individuals were asked to pick 'the single best response'
to dozens of questions, but often indicted more than one response.   If this
same questionnaire were administered electronically, the responses could
have been been constrained, but sadly, this was not the case.  

## Alternatives.  

Before describing what multi-factors are and how they work, there are
several obvious alternatives that may suffice in some cases, including (1) a
ragged array, (2) an indicator matrix, (3) a data.frame with indicator
variables, or a character vector with delimited.  

To begin with an example, lets say we ask 3 individuals, to answer the
quesiton "How do you feel about the Seattle Seahawks?  (a) I don't like
them, (b) I love them, or (c) I've never heard of the Seattle Seahawks".
furthermore lets assume that the first respondent replied 'a', the second
respondent replied 'b', and the third respondent replid 'b' and 'c'.

The following are three solutions to modeling these responses, each of which
may be preferred in some situations:


```{r}
# A ragged array
(response_list <- list('a','b',c('b','c')))

# an indicator matirx
(response_matrix <- matrix(c(T,F,F,
						     F,T,F,
						     F,T,T),
						     nrow=3,
						     byrow=T,
						     dimnames = list(1:3,
											 c('a','b','c'))))

# a data.frame with indicator variables. 
(response_df <- data.frame(a = c(T,F,F),
						   b = c(F,T,F),
						   c = c(F,T,T)))

# A comma delimited set of strings
(response_delim <- c('a','b','b,c'))

```

## About Multi-Factors (`mfactors`)

### Creating `mfactor` variables

Creating `mfactor` variables is similar to creating ordinary factors from a
character vector, except that the the argument `split` may be supplied
(which is passed to `base::strsplit`) and is used to split each string into
it's individual elements, as follows:

```{r}
library(mfactor)
mf <- mfactor(x =c(NA,"",'1','2','3','1,2','2,3','1,2,3')
	,split=','
	,levels = c('1','2','3','4')
	,labels = c('Never','Rarely','Sometimes','Often')
	,ordered = TRUE)
mf
```

Notice that several things happend in this call to mfactor. 

- The elements of the argument `x` were separated using the regular
  expression supplied to the argument `split`
- The resulting elements were checked against the expected values supplied
  in the `levels` argument, and the user would be warned if additonal values
  were observed in `x` (after splitting `x` of course)
- The values were re-labeled with labels supplied to the argument `labels`

At each step many common errors (e.g. invalid values in the argumen `x`) are
checked, and warnings or error issued in response, which can help with the
data cleaning process.  

Many other types of objects can also be  coerced to multi-factors via
mfactors, as in:

```{r}
# A ragged array
mfactor(response_delim,
		split = ',',
		levels=c('a','b','c'),
		labels=c("Love'em","Hate'em","Who?"))
mfactor(response_list,
		levels=c('a','b','c'),
		labels=c("Love'em","Hate'em","Who?"))
# Note that by default the levels are taken from the column names
# for data.frames and matricies (if they exist)
mfactor(response_matrix,
		labels=c("Love'em","Hate'em","Who?"))
mfactor(response_df,
		labels=c("Love'em","Hate'em","Who?"))
```

### Coersion to Ordinary Factors

Typically the goal of data cleaning is to remove anamalies like the ones
previously discucssed, so mfactrs are able to be converted to orinary
factors using 'as.factor'. 

```{r}
#conversion to factor types
as.factor(mf)
as.factor(mf,sep = '|')
```

The above conversions were not very useful, since, except for the fact that
the codes were exchanged for labels, the data are in the same shape they
were at the beining of this exercise.

The real power of multi-factors comes from the fact that we can supply an
*aggregating function*, which returns a single value when supplied a vector
input. Examples of aggregating functions include `min`,`max`, and `median`.  

```{r}
# The 'smallest' value from each element:
as.factor(mf,min) # or equivelently: min(mf)
# The 'largest' value from each element:
as.factor(mf,max) # or equivelently: max(mf)
# this median value provided (missing if even number of values provided)
as.factor(mf,median) 
```

Did you notice that in the last example, supplying `median` as the
aggregaing function introduced meissing values in `mf` and a warning stating
`Warning in as.factor.mfactor(mf, median): NAs introduced by coersion from
levels:"Never;Rarely", "Rarely;Sometimes"`?  This happed  because
aggregaging functions *should* return an integer between zero and
`length(levels(mf))` or a mising value), but in this case median returned a
non integer value.  Because the `median` of the selected values is often
desirable, the `mfactor` package includes 3 aggregating functions which
avoid this problem.

```{r}
# the median value in the absence of ties, and the upper value otherwise
as.factor(mf,upper.median)

# the median value in the absence of ties, and the lower median otherwise
as.factor(mf,lower.median)

# the median value in the absence of ties, otherwise a random selection from 
# the values closest to the arethmetic median.
as.factor(mf,median.random)
```

User defined aggregating functions can also be used

```{r}
AGG <- function(x)floor(median(x)[1])
as.factor(mf,AGG)
```

or alternatively, a string indicating the value to be assigned to each
mutipliy-valued element can be specified

```{r}
as.factor(mf,collapse='<Multiple-Values>')
```

### Setting Coersion Options Globally 

Default values for the arguments `collapse`, `sep`, and `none` in calls to
`as.factor` and `as.ordered` can also be set golbally as follows:

```{r}
options(mfactor.sep="::")
as.factor(mf)
options(mfactor.none= "<<<EMPTY>>>")
as.factor(mf)
options(mfactor.collapse="<<<Various>>>")
as.factor(mf)
options(mfactor.collapse=median.random)
as.factor(mf)
```

Note that arguments supplied to `as.factor()` will always override global
options.

```{r,echo=FALSE}
options(mfactor.sep=NULL)
options(mfactor.none= NULL)
options(mfactor.collapse=NULL)
```

## The 'NONE" type

As mentioned earlier, each element of an multi-factor can be
thought of as the set of responses to a question indicating "Check all that
apply'.  In the case that none of the options apply (and there is no option
for 'None'), an element containing no values ("None of these apply to me")
is distinct from a missing value ("User was not asked this question"). 

There is a special `NONE` object which represents the former concept("None 
of these apply to me"), which can be used to test for elements in which no
response is selected:

```{r}
mf == NONE
```

`is.none` and `is.none<-` can also be used to query and assign `NONE`
values:

```{r}
is.none(mf)
x <- mf
is.none(x[1:5]) <- T
x
```
and the string used to indicate 'NONE' values can be set globally 

```{r}
options('mfactor.none'='<Hello World>') 
x
```

```{r,echo=F}
options('mfactor.none'='<None>') # glabally
rm(x)
```

## Conversion to other types

Multi-factors can be coerced to a variety of other types as follows:

```{r}
# Ragged Arrays:
as.list(mf)
(mf_ragged_array <- as.list(mf))

# Character Vectors
(mf_character <- as.character(mf,sep='|'))

# Indicator Matricies.
(mf_matrix <- as.matrix(mf))
```

As a bit of syntactic sugar, a single column of the indicator matrix can be
extracted using the `$` operator 

```{r}
mf$Sometimes
```

## Algebraic Operators

The operators `+` and `-` are defined for multifactors that share a common
set of `levels()`, which can be thought of as `Vectorize`d versions `union`
and `setdiff`:

```{r}

(x <- mfactor(letters[1:6],levels = letters[1:10])) #length = 6
(y <- mfactor(letters[8:3],levels = letters[1:10])) #length = 6
(q <- mfactor(letters[1:3],levels = letters[1:10])) #length = 3
(p <- mfactor(letters[1:5],levels = letters[1:10])) #length = 5

x + y

# note that the usual recycling rules apply
x + q 

x + p # Warning: longer object length is not a multiple of shorter object length
```

Type coercion is automatically applied by the  `+` and `-` operators:

```{r}
x - 'a'
x + list('a','b') # notice the recycling rule is used because list('a','b') has length > 1
x + list(c('a','b')) # now 'a' and 'b' are added to each element...
```

```{r,echo=FALSE}
rm(x,y,p,q)
```


## Comparisons operators

Comparisons are available for ordered multi-factors however, these can be
somewhat non-intuitive.  There equality 

```{r}
mf2 <- mf
mf <  mf2 # implemented as max(mf) <  min(mf2)
mf <= mf2 # implemented as max(mf) <= min(mf2)
mf >  mf2 # implemented as min(mf) >  max(mf2)
mf >= mf2 # implemented as min(mf) >= max(mf2)
```

Note that the options `mfactor.strict.compare` affects how the equaliti
operators evalute the comparison:

```{r}
# the default behavior
options( mfactor.strict.compare=TRUE) 
mf == mf2 # essentially a vectorized version of `identical(a,b)`
mf != mf2 # essentially a vectorized version of `!identical(a,b)`

# the alternate behavior
options( mfactor.strict.compare=FALSE)
mf == mf2 # essentially a vectorized version of `any(a %in%  b)`
mf != mf2 # essentially a vectorized version of`!any(a %in%  b)`
```

Lastly, there is implied type coercion for the comparison operators
```{r}
mf < 'Sometimes'
mf <= 'Sometimes'
mf > 'Sometimes'
mf >= 'Sometimes'
```

### Subsets and Subset Assignment 

Multi-factors behave in a similar way to ordinary factors in terms of
subsets and subset assingnment

```{r}
mf[[2]] 
mf[2:4]
mf[3] <- 'Sometimes'
mf
```
except that we can assign a list containing subsets of `levels(x)` when
assiging values to a subset of a multi-factor.

```{r}
# subset assignment
mf[[8]] <- list(c('Sometimes','Often')) # a new multi factor level
mf

# if a list is assigned, recycling happens at the list-element level
mf[1:2] <- list(c('Never','Often'))
mf

# if a vector is assigned, recycling happens at the vector-element level
mf[6:10] <- c('Sometimes','Often')
mf

# invalid factor level assignment creates missing values and issues a warning
mf[9:10] <- c('hello','world')
```

## Miscelaneous

Like other vectors, multi-factors can be extended or truncated using the
`length<-`, missing values can be set queried and set via `is.na` and
`is.na<-`, and the names can be set and queried via `names` and `names<-`:

```{r}
# extension or truncation
length(mf)
length(mf) <- 20
length(mf)
length(mf) <- 5
length(mf)

# setting and identifying missing values
is.na(mf)
is.na(mf) <- 5
is.na(mf)

# element names
names(mf) <- paste0('item',seq_along(mf))
mf
```

And like factors, the levels can be queried and reassigned via `levels` and
`levels<-`

```{r}
levels(mf)
levels(mf) <- LETTERS[15:18]
levels(mf)
```



